Estimating duration in the presence of misclassification

### 2024-02-29 transmission.tex
edit backcalc methods

I keep the same logistic regression model in terms of prevalence, transforming its posterior distribution of prevalence to a posterior distribution of incidence.
This results in a two-step process: first, the same inference process described in \cref{E-biology-data:sec:MRP} is used to estimate prevalence; second, the prevalence is transformed to incidence.
An alternative would be an inference procedure that estimates incidence proportion, fit directly to the RT-PCR test results.
This is required to fulfil several desirable features of the model.
Denote the quantity to be inferred (which could be either incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
The strata will represent at least age group and region, although including ethnicity and sex is also due to non-response in the survey (see \cref{E-intro:sec:cis-strata}).
For many combinations of stratum $i$ and day $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
The statistical model and inference procedure should have the following properties.
\begin{enumerate}
    \item The smoothing model over time should penalise changes from continuous exponential growth.
        Penalising changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:cis}).
    \item Borrow information across the different covariates.
        The borrowing should penalise towards the distribution of incidence/prevalence across strata in the same geographic region being constant.
        Standard stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) predict this, if behavioural changes are independent of strata and there is negligible impact of immunity.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see\todo{ref section on SRS computational limitations}).
\end{enumerate}

Properties 1 and 2 suggest a model of the form $\logit(\pi_{it}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalised (\ie changes away from linear are penalised).
The $\logit$ link function both limits $\pi_{it}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{it}$~\autocite{ealesAppropriately}.
Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalised.
To add additional flexibility,

@@

Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
In particular, using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
Instead, previous MRP implementations have used the Integrated Nested Laplace Approximation (INLA)~\autocite{rueINLA}, implemented in R-INLA~\autocite{RINLA} (see \cref{transmission:sec:INLA}).
theta

\todo[inline]{Is it OK that $x$ and $y$ in the following two paragraphs don't align with the rest of the thesis chapter? Should I use alterantive notation - but then different to standard in the INLA papers.}
INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models})~\autocite{rueINLAseminar,rueINLA,martinoINLAtutorial}.
An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $p(\vec{y} \mid \vec{\varphi}, \vec{x}) = \prod_j p(y_j \mid \eta_j(\vec{x}), \vec{\varphi})$.

In this context, each element $\vec{y}$ corresponds to the observed RT-PCR results in one stratum on one day.
Hence, $\eta_{j}$ must determine the distribution of $y_{j}$ (up to some hyperparameters); where $j$ can be mapped to a mapping of a stratum $i$ and day $t$.
In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
If $\eta_j = \logit(\pi_{it})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is a LGM.
However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
In particular, if $\eta_j$ is the prevalence, then $\eta_{it} = \sum_{k=-\infty}^0 \pi_{ik} S(k) = \sum_{k=-\infty}^0 \expit(g(i, k)) S(k)$.
As $g(i, k)$ is linear in the model parameters, $\vec{x}$, then $\eta_{it}$ is not, violating the requirements of a LGM.
Therefore, the application of the INLA approximation, while satisfying the other desirable properties, requires that the regression model predicts RT-PCR prevalence, not incidence proportion.

>>>>

% I keep the same logistic regression model for predicting prevalence, then transform its posterior distribution of prevalence to a posterior distribution of incidence.
This results in a two-step process: first, the same inference process described in \cref{E-biology-data:sec:MRP} is used to estimate prevalence; second, the prevalence is transformed to incidence.
This is required to fulfil several desirable features of the model.

An alternative would be an inference procedure that estimates incidence proportion, fit directly to the RT-PCR test results.
Denote the quantity to be inferred (which could be either incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
The strata will represent at least age group and region, although including ethnicity and sex is also due to non-response in the survey (see \cref{E-intro:sec:cis-strata}).
For many combinations of stratum $i$ and day $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
The statistical model and inference procedure should have the following properties.
\begin{enumerate}
    \item The smoothing model over time should penalise changes from continuous exponential growth.
        Penalising changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:cis}).
    \item Borrow information across the different covariates.
        The borrowing should penalise away from changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
        Standard stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) predict this, if behavioural changes are independent of strata and there is negligible impact of immunity.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

Properties 1 and 2 suggest a model of the form $\logit(\pi_{it}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalised (\ie changes away from linear are penalised).
The $\logit$ link function both limits $\pi_{it}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{it}$~\autocite{ealesAppropriately}.
Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalised.

@@

Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
In particular, using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
Instead, previous MRP implementations used INLA (see \cref{E-biology-data:sec:MRP}).

\todo[inline]{Is it OK that $x$ and $y$ in the following two paragraphs don't align with the rest of the thesis chapter? Should I use alterantive notation - but then different to standard in the INLA papers.}
INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models})~\autocite{rueINLAseminar,rueINLA,martinoINLAtutorial}; see \cref{E-transmission:sec:INLA} for further details on INLA.
An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $p(\vec{y} \mid \vec{\varphi}, \vec{x}) = \prod_j p(y_j \mid \eta_j(\vec{x}), \vec{\varphi})$.

In this context, each $y_j$ corresponds to the observed RT-PCR results in one stratum on one day.
Hence, $\eta_{j}$ must determine the distribution of $y_{j}$ (up to some hyperparameters); where $j$ uniquely identifies a stratum $i$ and day $t$.
In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
If $\eta_j = \logit(\pi_{it})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is a LGM.
However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
In particular, if $\eta_j$ is the prevalence, then $\eta_{it} = \sum_{k=-\infty}^0 \pi_{ik} S(k) = \sum_{k=-\infty}^0 \expit(g(i, k)) S(k)$.
As $g(i, k)$ is linear in the model parameters, $\vec{x}$, then $\eta_{it}$ is not, violating the requirements of a LGM.
Therefore, the application of the INLA approximation, while satisfying the other desirable properties, requires that the regression model predicts RT-PCR prevalence, not incidence proportion.
